---
title: "FinalProject Data 205 file 1"
author: "Miriam Schoenbaum"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

# Load packages

```{r load-packages, message=FALSE}
library(tidyverse)
library(data.table)
library(naniar)
library(lubridate)
library(chron)
library(generalhoslem)
library(performance)
```

# Check timezone

```{r}
Sys.timezone()
```

# Load data
```{r}
# Data Montgomery
nonmotorist <- fread('https://data.montgomerycountymd.gov/resource/n7fk-dce5.csv?$limit=10000')
incident <- fread('https://data.montgomerycountymd.gov/resource/bhju-22kf.csv?$limit=150000')
```

```{r}
# Maryland State Police
# Data available at https://mdsp.maryland.gov/Pages/Dashboards/CrashDataDownload.aspx
# Steps for getting the file:
# Selection parameters: Crash Date Begin 1/1/2018, County of Crash Montgomery, remainder on default
# Download data
# NONMOTORIST data
# Include default columns (ReportNumber, NonMotoristID) plus NonMotoristDOB and NonMotoristSex
# save in working directory as the default file download name "CrashMap_NONMOTORIST_data.csv"
setwd("C:/Data_205/project_files")
MSP <- read.csv('CrashMap_NONMOTORIST_data.csv')
```

# Wrangle the MSP data

```{r}
# Change non-motorist sex from character to factor & convert unknown to NA
MSP$NonMotoristSex <- as.factor(MSP$NonMotoristSex)
MSP <- MSP |>
  mutate(NonMotoristSex = case_when(
    NonMotoristSex == "U" ~ NA,
    NonMotoristSex == "M" ~ "M",
    .default = "F"
  ))
table(MSP$NonMotoristSex)
# Change non-motorist date of birth from character to date
MSP$NonMotoristDOB <- as.Date(MSP$NonMotoristDOB,"%m/%d/%Y")
# Change small letters in NonmotoristID to capital letters
MSP$NonMotoristID <- toupper(MSP$NonMotoristID)
```

# Wrangle the nonmotorist data

```{r}
# Change a lot of variables from character to factor and order factors as needed
nonmotorist$agency_name <- as.factor(nonmotorist$agency_name)
nonmotorist$acrs_report_type <- as.factor(nonmotorist$acrs_report_type)
nonmotorist$route_type <- as.factor(nonmotorist$route_type)
nonmotorist$cross_street_type <- as.factor(nonmotorist$cross_street_type)
nonmotorist$municipality <- as.factor(nonmotorist$municipality)
nonmotorist$related_non_motorist <- as.factor(nonmotorist$related_non_motorist)
nonmotorist$collision_type <- as.factor(nonmotorist$collision_type)
nonmotorist$weather <- as.factor(nonmotorist$weather)
nonmotorist$surface_condition <- as.factor(nonmotorist$surface_condition)
nonmotorist$light <- as.factor(nonmotorist$light)
nonmotorist$traffic_control <- as.factor(nonmotorist$traffic_control)
nonmotorist$pedestrian_type <- as.factor(nonmotorist$pedestrian_type)
nonmotorist$pedestrian_movement <- as.factor(nonmotorist$pedestrian_movement)
nonmotorist$pedestrian_actions <- as.factor(nonmotorist$pedestrian_actions)
nonmotorist$pedestrian_location <- as.factor(nonmotorist$pedestrian_location)
nonmotorist$pedestrian_obeyed_traffic_signal <- as.factor(nonmotorist$pedestrian_obeyed_traffic_signal)
nonmotorist$pedestrian_visibility <- as.factor(nonmotorist$pedestrian_visibility)
nonmotorist$at_fault <- as.factor(nonmotorist$at_fault)
nonmotorist$injury_severity <- as.factor(nonmotorist$injury_severity)
nonmotorist$injury_severity <- factor(nonmotorist$injury_severity, ordered = TRUE, levels = c("NO APPARENT INJURY", "POSSIBLE INJURY", "SUSPECTED MINOR INJURY", "SUSPECTED SERIOUS INJURY", "FATAL INJURY"))
# Create date variable for calculating age
nonmotorist <- nonmotorist |>
  mutate(crashdate = as.Date(nonmotorist$crash_date_time,"%m/%d/%Y", tz = ''))
# Create year variable
nonmotorist <- nonmotorist |>
  mutate(year = as.numeric((format(nonmotorist$crashdate,"%Y"))))
# Create on/off road variable
nonmotorist <- nonmotorist |>
  replace_with_na(replace = list(off_road_description = "")) |>
  mutate(on_offroad = if_else(is.na(off_road_description), "On road", "Off road"))
nonmotorist$on_offroad <- as.factor(nonmotorist$on_offroad)
# Create serious/fatal injury variable
nonmotorist <- nonmotorist |>
  mutate(serious_fatal_injury = case_when(
    injury_severity == "FATAL INJURY" ~ "serious/fatal",
    injury_severity == "SUSPECTED SERIOUS INJURY" ~ "serious/fatal",
    .default = "not serious/fatal"
  ))
nonmotorist$serious_fatal_injury <- as.factor(nonmotorist$serious_fatal_injury)
# Create a day/night variable (day = 7 am to 9 pm, night = 9 pm to 7 am, based on the times for non-motorist crashes in school walk zones in the Safe Streets Act)
nonmotorist <- nonmotorist |>
  mutate(crashtime = format(as.POSIXct(crash_date_time),format = "%H:%M:%S"))
nonmotorist$crashtime <- times(nonmotorist$crashtime)
nonmotorist <- nonmotorist |>
  mutate(day_night = case_when(
    crashtime >= "07:00:00" & crashtime < "21:00:00" ~ "day",
    .default = "night"
  ))
nonmotorist$day_night <- as.factor(nonmotorist$day_night)
# Create month variable and seasonality variable
nonmotorist <- nonmotorist |>
  mutate(month = month(crash_date_time))
nonmotorist <- nonmotorist |> 
  mutate(season = case_when (
    month == "12" | month == "1" | month == "2" ~ "winter",
    month == "3" | month == "4" | month == "5" ~ "spring",
    month == "6" | month == "7" | month == "8" ~ "summer",
    .default = "fall"
  ))
nonmotorist$season <- factor(nonmotorist$season, ordered = TRUE, levels = c("winter", "spring", "summer", "fall"))
# Create "dart dash" variable
nonmotorist <- nonmotorist |>
  mutate(dartdash = if_else(pedestrian_actions == "DART DASH", "dart dash", "not dart dash"))
nonmotorist$dartdash <- as.factor(nonmotorist$dartdash)
# Create pedestrian/non-pedestrian variable
nonmotorist <- nonmotorist |>
  mutate(pedestrian_yes_no = ifelse(pedestrian_type == "PEDESTRIAN", "pedestrian", "small wheeled thing"))
nonmotorist$pedestrian_yes_no <- as.factor(nonmotorist$pedestrian_yes_no)
# Create binary at-fault variable
nonmotorist <- nonmotorist |>
  mutate(at_fault.binary = ifelse(at_fault == "No", "Not at fault", "At fault or unknown"))
nonmotorist$at_fault.binary <- as.factor(nonmotorist$at_fault.binary)
# Create variables for day of the week and weekday/weekend
nonmotorist <- nonmotorist |>
  mutate(weekday.day = weekdays.POSIXt(crash_date_time)) |>
  mutate(weekdays_y_n = case_when(
    weekday.day == "Sunday" ~ "weekend",
    weekday.day == "Saturday" ~ "weekend",
    .default = "weekday"))
nonmotorist$weekday.day <- factor(nonmotorist$weekday.day, ordered = TRUE, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
nonmotorist$weekdays_y_n <- factor(nonmotorist$weekdays_y_n)
# Create variable for county road
nonmotorist <- nonmotorist |>
  mutate(county_road = if_else(route_type == "County" | cross_street_type == "County", "County", "Not county"))
nonmotorist$county_road = as.factor(nonmotorist$county_road)
```


# Wrangle the incident data

```{r}
# Change chosen variables from character to factor
incident$hit_run <- as.factor(incident$hit_run)
incident$nontraffic <- as.factor(incident$nontraffic)
incident$junction <- as.factor(incident$junction)
```

# Start the main data table by adding the incident fields to the non-motorist table

```{r}
# Join tables
datatable <- left_join(nonmotorist, incident, by = join_by(report_number == report_number), relationship = "many-to-one")
# drop duplicate variables
datatable <- datatable |>
  dplyr::select(-ends_with(".y")) 
```

# Add the MSP fields to the main data table

```{r}
datatable <- left_join(datatable, MSP, by = join_by(person_id == NonMotoristID))
```

# Drop 2015, 2016, 2017 from the main data table because the MSP data start in 2018

```{r}
datatable <- subset(datatable, year > 2017)
```

# Create age field

```{r}
datatable <- datatable |>
  mutate(age = (crashdate - NonMotoristDOB)/365)
datatable$age <- as.integer(datatable$age)
```

# Create 2 age category fields

```{r}
# Create binary age variable
datatable <- datatable |>
  mutate(adult_yes_no = if_else((age < 19), "Not adult", "Adult"))
datatable$adult_yes_no <- as.factor(datatable$adult_yes_no)
# Create age group field
datatable <- datatable |>
  mutate(agegroup = case_when(
    age >= 0 & age < 5 ~ "0 to 4",
    age >= 5 & age <12 ~ "5 to 11",
    age >= 12 & age <15 ~ "12 to 14",
    age >= 15 & age <19 ~ "15 to 18",
    .default = "19+"
  ))
datatable$agegroup <- as.factor(datatable$agegroup)
datatable$agegroup <- factor(datatable$agegroup, ordered = TRUE, levels = c("0 to 4", "5 to 11", "12 to 14", "15 to 18", "19+"))
```

# Look at age group and adult/non-adult

```{r}
table(datatable$agegroup)
table(datatable$adult_yes_no)
hist(datatable$age, main = "Age (years) of non-motorists in police-reported crashes", xlab = "Years", ylab = "Number of non-motorists since 2018", col = "lightblue", breaks = 120) 
pie(table(datatable$agegroup), main = "Non-motorists by age group in police-reported crashes", col = c("white", "lightblue", "blue", "darkblue", "red"))
```

# Time variables

```{r}
table(datatable$year, datatable$agegroup)
table(datatable$year, datatable$adult_yes_no)
table(datatable$month, datatable$agegroup)
table(datatable$month, datatable$adult_yes_no)
table(datatable$season, datatable$agegroup)
table(datatable$season, datatable$adult_yes_no)
table(datatable$day_night, datatable$agegroup)
table(datatable$weekday.day, datatable$agegroup)
# Odds ratio for day or night
day_night_table <- table(datatable$day_night, datatable$adult_yes_no)
day_night_table
mosaicplot(day_night_table, color = c("red", "blue"), xlab ="Day or night", ylab = "Adult or not adult", main = "Adult/Not adult and Day/Night")
fisher.test(day_night_table)
# Odds ratio for weekday or weekend
weekday_table<- table(datatable$weekdays_y_n, datatable$adult_yes_no)
weekday_table
mosaicplot(weekday_table, color = c("red", "blue"), xlab = "Weekday or weekend", ylab = "Adult or not adult", main = "Adult/Not adult and Weekday/Weekend")
fisher.test(weekday_table)
```

# Charts for time

```{r}
datatable$month_year <- format(as.Date(datatable$crash_date_time.x), "%Y-%m")
seasonal_age <- datatable |>
  count(month_year, agegroup)
# by age group
plot1 <- ggplot(seasonal_age, aes(x = month_year, y = n, group = agegroup, color = agegroup)) +
 # make it a line plot
  geom_line(linewidth = 0.5) +
 # add points
  geom_point() +
  #set the x axis scale and labels
  scale_x_discrete(breaks = c("2018-01", "2019-01", "2020-01", "2021-01", "2022-01", "2023-01")) +
  #set the title
  ggtitle(label = "Non-motorists with police crash reports, by age group", subtitle = "By month, January 2018-present") +
  # set the label for the x axis
  xlab("Year-Month") +
# set the label for the y axis
  ylab("Number of non-motorists") +
# set the label for the legend
  labs(color = "Age group") + 
  # set the colors for the age groups 
  scale_color_manual(values = c("dodgerblue4", "dodgerblue3", "dodgerblue2", "dodgerblue", "firebrick1"))
plot1
# by age group
seasonal_age <- datatable |>
  count(month_year, agegroup)
plot1 <- ggplot(seasonal_age, aes(x = month_year, y = n, group = agegroup, color = agegroup)) +
 # make it a line plot
  geom_line(linewidth = 0.5) +
 # add points
  geom_point() +
  #set the x axis scale and labels
  scale_x_discrete(breaks = c("2018-01", "2019-01", "2020-01", "2021-01", "2022-01", "2023-01")) +
  #set the title
  ggtitle(label = "Non-motorists with police crash reports, by age group", subtitle = "By month, January 2018-present") +
  # set the label for the x axis
  xlab("Year-Month") +
# set the label for the y axis
  ylab("Number of non-motorists") +
# set the label for the legend
  labs(color = "Age group") + 
  # set the colors for the age groups 
  scale_color_manual(values = c("dodgerblue4", "dodgerblue3", "dodgerblue2", "dodgerblue", "firebrick1"))
plot1
# by adult/non-adult
seasonal_adult_nonadult <- datatable |>
  count(month_year, adult_yes_no)
seasonal_adult_nonadult <- na.omit(seasonal_adult_nonadult)
plot2 <- ggplot(seasonal_adult_nonadult, aes(x = month_year, y = n, group = adult_yes_no, color = adult_yes_no)) +
 # make it a line plot
  geom_line(linewidth = 0.5) +
 # add points
  geom_point() +
  #set the x axis scale and labels
  scale_x_discrete(breaks = c("2018-01", "2019-01", "2020-01", "2021-01", "2022-01", "2023-01")) +
  #set the title
  ggtitle(label = "Non-motorists with police crash reports, by adult (19+) or non-adult", subtitle = "By month, January 2018-present") +
  # set the label for the x axis
  xlab("Year-Month") +
# set the label for the y axis
  ylab("Number of non-motorists") +
# set the label for the legend
  labs(color = "Adult or non-adult") +
  # set the colors for the age groups 
  scale_color_manual(values = c("firebrick1", "dodgerblue4"))
plot2
# without adults
kids_seasonal_age <- subset(seasonal_age, agegroup != "19+")
plot3 <- ggplot(kids_seasonal_age, aes(x = month_year, y = n, group = agegroup, color = agegroup)) +
 # make it a line plot
  geom_line(linewidth = 0.5) +
 # add points
  geom_point() +
  #set the x axis scale and labels
  scale_x_discrete(breaks = c("2018-01", "2019-01", "2020-01", "2021-01", "2022-01", "2023-01")) +
  #set the title
  ggtitle(label = "Non-motorists with police crash reports, by age group for non-adults (<19)", subtitle = "By month, January 2018-present") +
  # set the label for the x axis
  xlab("Year-Month") +
# set the label for the y axis
  ylab("Number of non-motorists") +
# set the label for the legend
  labs(color = "Age group for non-adults") +
  # set the colors for the age groups 
  scale_color_manual(values = c("skyblue", "blue", "red", "darkred"))
plot3
```

# Roads variables

```{r}
table(datatable$number_of_lanes, datatable$agegroup)
table(datatable$number_of_lanes, datatable$adult_yes_no)
table(datatable$on_offroad, datatable$agegroup)
on_offroad_table <- table(datatable$on_offroad, datatable$adult_yes_no)
on_offroad_table
# Odds ratio for on or off road
mosaicplot(on_offroad_table, color = c("red", "blue"), xlab ="On or off road", ylab = "Adult or not adult", main = "Adult/Not adult and On road/Off road")
fisher.test(on_offroad_table)
#Odds ratio for county or non county road
countyroad_table <- table(datatable$county_road, datatable$adult_yes_no)
countyroad_table
mosaicplot(countyroad_table, color = c("red", "blue"), xlab ="County or non county road", ylab = "Adult or not adult", main = "Adult/Not adult and County or non county road")
fisher.test(countyroad_table)
```

# Crash variables

```{r}
table(datatable$serious_fatal_injury, datatable$agegroup)
table(datatable$at_fault.binary, datatable$agegroup)
table(datatable$dartdash, datatable$agegroup)
table(datatable$agency_name.x, datatable$agegroup)
table(datatable$agency_name.x, datatable$adult_yes_no)
table(datatable$agency_name.x, datatable$at_fault.x)
table(datatable$hit_run, datatable$agegroup)
# Odds ratio for serious or fatal injury
serious_fatal_table <- table(datatable$serious_fatal_injury, datatable$adult_yes_no)
serious_fatal_table
mosaicplot(serious_fatal_table, color = c("red", "blue"), xlab ="Serious/fatal or not serious/fatal injury", ylab = "Adult or not adult", main = "Adult/Not adult and Serious/fatal or Not serious/fatal injury")
fisher.test((serious_fatal_table))
# Odds ratio for at fault
at_fault_table <- table(datatable$at_fault.binary, datatable$adult_yes_no)
at_fault_table
mosaicplot(at_fault_table, color = c("red", "blue"), xlab ="At fault or unknown/Not at fault", ylab = "Not adult/Adult", main = "Not adult/Adult and At fault or unknown/Not at fault")
fisher.test(at_fault_table)
# Odds ratio for "dart dash"
dartdash_table <- table(datatable$dartdash, datatable$adult_yes_no)
dartdash_table
mosaicplot(dartdash_table, color = c("red", "blue"), xlab ="'Dart dash' or other action", ylab = "Adult or not adult", main = "Adult/Not adult and 'Dart dash'/Other action")
fisher.test(dartdash_table)
# Odds ratio for hit & run
hit_run_table <- table(datatable$hit_run, datatable$adult_yes_no)
hit_run_table
mosaicplot(hit_run_table, color = c("red", "blue"), xlab ="Hit & run or not hit & run", ylab = "Adult or not adult", main = "Adult/Not adult and Hit & run/Not hit & run")
fisher.test(hit_run_table)
```

# Extra analysis for non-motorists at fault who were hit in a sidewalk or crosswalk and didn't disobey the signal
```{r}
# Create dataset for in crosswalk or on sidewalk 
crossorsidewalk <- filter(datatable, pedestrian_location == "AT INTERSECTION BUT NO CROSSWALK" | pedestrian_location == "AT INTERSECTION MARKED CROSSWALK" | pedestrian_location == "DRIVEWAY ACCESS MARKED CROSSWALK" | pedestrian_location == "MIDBLOCK MARKED CROSSWALK" | pedestrian_location == "ON ROADWAY AT CROSSWALK" | pedestrian_location == "SHARED USE PATH OR TRAILS" | pedestrian_location == "SIDEWALK")
crossorsidewalk_table <- table(crossorsidewalk$at_fault.binary, crossorsidewalk$adult_yes_no)
crossorsidewalk_table
mosaicplot(crossorsidewalk_table, color = c("red", "blue"), xlab ="At fault or unknown/Not at fault", ylab = "Not adult/Adult", main = "For nonmotorists hit in crosswalks or on sidewalks:\n Not adult/Adult and At fault or unknown/Not at fault")
fisher.test(crossorsidewalk_table)
crosswalk_did_not_disobey <- filter(crossorsidewalk, pedestrian_obeyed_traffic_signal != "DISOBEYED PED. SIGNAL")
didnotdisobeytable <- table(crosswalk_did_not_disobey$at_fault.binary, crosswalk_did_not_disobey$adult_yes_no)
didnotdisobeytable
mosaicplot(didnotdisobeytable, color = c("red", "blue"), xlab ="At fault or unknown/Not at fault", ylab = "Not adult/Adult", main = "For nonmotorists hit in crosswalks or on sidewalks, \n who did not disobey a signal: \n Not adult/Adult and At fault or unknown/Not at fault")
fisher.test(didnotdisobeytable)
```


# Transportation mode variables

```{r}
table(datatable$pedestrian_type, datatable$agegroup)
table(datatable$pedestrian_type, datatable$adult_yes_no)
table(datatable$pedestrian_yes_no, datatable$agegroup)
pedestrian_table <- table(datatable$pedestrian_yes_no, datatable$adult_yes_no)
pedestrian_table
mosaicplot(pedestrian_table, color = c("red", "blue"), xlab ="Pedestrian or Small thing with wheels", ylab = "Adult or not adult", main = "Adult/Not adult and Pedestrian/Small thing with wheels")
fisher.test(pedestrian_table)
```

# Gender (wrangled to binary) variables
```{r}
table(datatable$NonMotoristSex, datatable$agegroup)
table(datatable$NonMotoristSex, datatable$adult_yes_no)
gender_table <- table(datatable$NonMotoristSex, datatable$adult_yes_no)
mosaicplot(gender_table, color = c("red", "blue"), xlab ="Male or Female", ylab = "Adult or not adult", main = "Adult/Not adult and Male or Female")
fisher.test(gender_table)
```


# Chi squared tests to assess significant relationships

```{r}
# Time
# Year - adult yes/no
chisq.test(datatable$year, datatable$adult_yes_no)
# Month - adult yes/no
chisq.test(datatable$month, datatable$adult_yes_no)
# Season - adult yes/no
chisq.test(datatable$season, datatable$adult_yes_no)
# Day/night - adult yes/no
chisq.test(datatable$day_night, datatable$adult_yes_no)
# Weekday/weekend - adult yes/no
chisq.test(datatable$weekdays_y_n, datatable$adult_yes_no)
# Road
# On or off road - adult yes/no
chisq.test(datatable$on_offroad, datatable$adult_yes_no)
# County or non county road - adult yes/no
chisq.test(datatable$county_road, datatable$adult_yes_no)
# Crash
# Serious or fatal
chisq.test(datatable$serious_fatal_injury, datatable$adult_yes_no)
# At fault
chisq.test(datatable$at_fault.x, datatable$adult_yes_no)
# Dart dash
chisq.test(datatable$dartdash, datatable$adult_yes_no)
# Mode
chisq.test(datatable$pedestrian_yes_no, datatable$adult_yes_no)
# Gender
chisq.test(datatable$NonMotoristSex, datatable$adult_yes_no)
```

# Logistic regression model for non-motorist at fault

```{r}
# Look at the variables separately for collinearity
table(datatable$at_fault.binary, datatable$day_night)
chisq.test(datatable$at_fault.binary, datatable$day_night)
table(datatable$at_fault.binary, datatable$on_offroad)
chisq.test(datatable$at_fault.binary, datatable$on_offroad)
table(datatable$at_fault.binary, datatable$county_road)
chisq.test(datatable$at_fault.binary, datatable$county_road)
table(datatable$at_fault.binary,datatable$pedestrian_yes_no)
chisq.test(datatable$at_fault.binary,datatable$pedestrian_yes_no)
table(datatable$at_fault.binary, datatable$adult_yes_no)
chisq.test(datatable$at_fault.binary, datatable$adult_yes_no)
table(datatable$at_fault.binary, datatable$serious_fatal_injury)
chisq.test(datatable$at_fault.binary, datatable$serious_fatal_injury)
table(datatable$at_fault.binary, datatable$hit_run)
chisq.test(datatable$at_fault.binary, datatable$hit_run)
table(datatable$at_fault.binary, datatable$NonMotoristSex)
chisq.test(datatable$at_fault.binary, datatable$NonMotoristSex)
table(datatable$at_fault.binary, datatable$weekdays_y_n)
chisq.test(datatable$at_fault.binary, datatable$weekdays_y_n)
# Model 1: with transportation mode (pedestrian yes/no)
Model <- glm(at_fault.binary ~ day_night + on_offroad + pedestrian_yes_no + adult_yes_no + serious_fatal_injury + hit_run + NonMotoristSex, data = datatable, family = binomial(link='logit'))
summary(Model)
confint(Model)
# Model 2: without transportation mode (pedestrian yes/no)
Model2 <- glm(at_fault.binary ~ day_night + on_offroad + adult_yes_no + serious_fatal_injury + hit_run + NonMotoristSex, data = datatable, family = binomial(link='logit'))
summary(Model2)
confint(Model2)
# Check for goodness of fit for both models
Modeldata <- model.frame(Model)
Model2data <- model.frame(Model2)
logitgof(Modeldata$at_fault.binary, fitted(Model))
logitgof(Model2data$at_fault.binary, fitted(Model2))
# Check for multicollinearity for both models
check_collinearity(Model)
check_collinearity(Model2)
```

# Make a file for just crashes on county roads
```{r}
countydatatable <- filter(datatable, county_road == "County")
```

# Export csv file for geospatial analysis in ArcGIS Pro
```{r}
write.csv(datatable, "C:/Data_205/project_files/GISfile/datatable.csv", row.names=FALSE)
```



